On the power of conditional independence testing \\ under model-X

### 2021-07-20 crtpower.tex
Updated abstract

\begin{abstract}
For testing conditional independence (CI) of a response $Y$ and a predictor $X$ given covariates $Z$, the recently introduced model-X (MX) framework has been the subject of active methodological research, especially in the context of MX knockoffs and their successful application to genome-wide association studies. In this paper, we study the power of MX CI tests, yielding quantitative explanations for empirically observed phenomena and novel insights to guide the design of MX methodology. We show that any valid MX CI test must also be valid conditionally on $\sry, \srz$; this conditioning allows us to reformulate the problem as testing a point null hypothesis involving the conditional distribution of $X$. The Neyman-Pearson lemma then implies that the conditional randomization test (CRT) based on a likelihood statistic is the most powerful MX CI test against a point alternative. We also obtain a related optimality result for MX knockoffs. We show that the CRT resampling distribution of an appropriately normalized test statistic converges to a standard normal distribution under the MX assumption, leading to an asymptotically equivalent test based on explicit critical values (and therefore no need for resampling) with uniform asymptotic type-I error control. In fact, this test is valid under the assumption that \textit{only the first two moments of $X$ given $Z$ are known}, a significant relaxation of MX. Finally, we derive expressions for the power of this test (and that of the asymptotically equivalent CRT) against local semiparametric alternatives, explicitly capturing the prediction error of the underlying learning algorithm. 
\end{abstract}

>>>>

\begin{abstract}
For testing conditional independence (CI) of a response $Y$ and a predictor $X$ given covariates $Z$, the recently introduced model-X (MX) framework has been the subject of active methodological research, especially in the context of MX knockoffs and their successful application to genome-wide association studies. In this paper, we study the power of MX CI tests, yielding quantitative explanations for empirically observed phenomena and novel insights to guide the design of MX methodology. We show that any valid MX CI test must also be valid conditionally on $\sry, \srz$; this conditioning allows us to reformulate the problem as testing a point null hypothesis involving the conditional distribution of $X$. The Neyman-Pearson lemma then implies that the conditional randomization test (CRT) based on a likelihood statistic is the most powerful MX CI test against a point alternative. We also obtain a related optimality result for MX knockoffs. Switching to an asymptotic framework with arbitrarily growing covariate dimension, we derive an expression for the limiting power of the CRT against local semiparametric alternatives in terms of the prediction error of the machine learning algorithm on which its test statistic is based. Finally, we exhibit a resampling-free test with uniform asymptotic Type-I error control under the assumption that \textit{only the first two moments of $X$ given $Z$ are known}, a significant relaxation of the MX assumption. 
\end{abstract}
